\textbf{Effective application of machine learning to microbiome based classification problems}

### 2019-10-14 submission/manuscript.tex
Jenna's edits last round

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage[table]{xcolor}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{caption}
\usepackage{hyperref}

\usepackage{helvet} % Helvetica font
\renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
\usepackage[T1]{fontenc}
\usepackage[labelfont=bf]{caption}

@@

\(\dagger\) To whom correspondence should be addressed:
\href{mailto:pschloss@umich.edu}{\nolinkurl{pschloss@umich.edu}}

@@

Machine learning (ML) modeling of the human microbiome has the potential
to identify microbial biomarkers and aid in diagnosis of many diseases
such as inflammatory bowel disease, diabetes, and colorectal cancer.
Progress has been made towards developing ML models that predict health
outcomes using bacterial abundances, but inconsistent adoption of
training and evaluation methods call the validity of these models into
question. Furthermore, there appears to be a preference by many
researchers to favor increased model complexity over interpretability.
To overcome these challenges, we trained seven models that used fecal
16S rRNA sequence data to predict the presence of colonic screen
relevant neoplasias (SRNs; n=490 patients, 261 controls and 229 cases).
We developed a generalizable pipeline to train, validate and interpret
the models. To show the effect of model selection, we assessed the
predictive performance, interpretability, and training time of
L2-regularized logistic regression, L1 and L2-regularized support vector
machines (SVM) with linear and radial basis function kernels, decision
trees, random forest, and extreme gradient boosted trees (XGBoost). The
random forest model performed best at detecting SRNs with an AUROC of
0.695 {[}IQR 0.651-0.739{]} but was slow to train (83.2 h) and difficult
to interpret. Despite its simplicity, L2-regularized logistic regression
followed random forest in predictive performance with an AUROC of 0.680
{[}IQR 0.625-0.735{]}, trained faster (12 min), and was more
interpretable. Our analysis showed that ML models should be chosen based
on the goal of the study, as the choice will inform expectations of
performance and interpretability.

@@

As the number of people represented in human microbiome datasets grow,
there is an increasing desire to use microbiome data to diagnose
diseases. However, the structure of the human microbiome is remarkably
variable among individuals to the point where it is often difficult to
identify the bacterial populations that are associated with diseases
using traditional statistical models. This variation is likely due to
the ability of many bacterial populations to fill the same niche such
that different populations cause the same disease in different
individuals. Furthermore, a growing number of studies have shown that it
is rare for a single bacterial species to be associated with a disease.
Instead, subsets of the microbiome account for differences in health.
Traditional statistical approaches do not adequately account for the
variation in the human microbiome and typically consider the protective
or risk effects of each bacterial population seperately (1). Recently,
machine learning models (ML) have grown in popularity among microbiome
researchers because we are able to sample large numbers of individuals
and because the models are effective at accounting for the interpersonal
microbiome variation and the ecology of the disease.

ML models can be used to increase our understanding of the variation in
the structure of existing data and in making predictions about new data.
Researchers have used ML models to diagnose and understand the
ecological basis of diseases such as liver cirrhosis, colorectal cancer,
inflammatory bowel diseases, obesity, and type 2 diabetes (2--19). The
task of diagnosing an individual relies on a rigorously validated model.
However, there are common methodological and reporting problems that
arise when applying ML to such data, that need to be addressed for the
field to progress. These problems include a lack of transparency in
which methods are used and how these methods are implemented; evaluating
models without a separate held-out test data; large variation between
the predictive performance on different folds of cross-validation; and
large variation between cross-validation and testing performances.
Nevertheless, the microbiome field is making progress to avoid some of
these pitfalls including validating their models on independent datasets
(8, 19, 20) and introducing ways to better use ML tools (21--24). More
work is needed to further improve reproducibility and minimize
overestimating for model performance.

Among microbiome researchers, the lack of justification when selecting a
modeling approach has been due to an implicit assumption that more
complex models are better. This has resulted in a trend towards using
models such as random forest and deep neural networks (3, 12, 25--27)
over simpler models such as logistic regression or other linear models
(19, 23, 28). Although in some cases complex models may capture
important non-linear relationships and therefore yield better
predictions, they result in black boxes that lack interpretability. Such
models require post hoc explanations to quantify the importance of each
feature in making predictions. Depending on the goal of the modeling,
researchers may choose to use different approaches. For example,
researchers trying to identify the microbiota associated with disease
may desire a more interpretable model, whereas clinicians may emphasize
predictive performance. Nonetheless, it is important to understand the
tradeoff and accept that the benefit of more complex, less interpretable
models may be minimal (29, 30). It is important for researchers to
justify their choice of modeling approach.

To showcase a rigorous ML pipeline and to shed light on how ML model
selection can affect modeling results, we performed an empirical
analysis comparing seven modeling approaches with the same dataset and
pipeline. We built three linear models with different forms of
regularization: L2-regularized logistic regression and L1 and
L2-regularized support vector machines (SVM) with a linear kernel. We
also trained four non-linear models: SVM with radial basis function
kernel, a decision tree, random forest and XGBoost. We compared their
predictive performance, interpretability, and training time. To
demonstrate the performance of these modeling approaches and our
pipeline, we used data from a previously published study that sought to
classify individuals as having normal colons or colonic lesions based on
the 16S rRNA gene sequences collected from fecal samples (4). This
dataset was selected because it is a relatively large collection of
individuals (N=490) connected to a clinically significant disease where
there is ample evidence that the disease is driven by variation in the
microbiome (2, 4, 5, 31). With this dataset we developed a ML pipeline
that can be used in many different scenarios for training and evaluating
models. This framework can be easily applied to other host-associated
and environmental microbiome datasets.

@@

First, we randomly split the data into training and test sets so that
the training set consisted of 80\% of the full dataset, while the test
set was composed of the remaining 20\% {[}Figure 1{]}. To maintain the
distribution of controls and cases found in the full dataset, we
performed stratified splits. For example, our full dataset included 490
individuals. Of these, 261 had normal colons (53\%) and 229 had a screen
relevant neoplasia (SRN; 46.7\%). A training set included 393
individuals, of which 209 had an SRN (53\%), while the test set was
composed of 97 individuals of which 52 had an SRN (54\%). The training
data was used to build and select the models and the test set was used
for evaluating the model.

We trained seven different models using the training data {[}Table 1{]}.
We focused on different classification algorithms and regularization
methods. Regularization helps to prevent overfitting by penalizing a
model that fits the training data too well. For regularized logistic
regression and SVM with a linear kernel, we used L2-regularization to
keep all potentially important features. For comparison, we also trained
an L1-regularized SVM with a linear kernel. L1-regularization on
microbiome data led to a sparser solution (i.e., forced many
coefficients to zero). To explore the potential for non-linear
relationships among features to improve classification, we trained
tree-based models including a decision tree, a random forest, and
XGBoost and an SVM with a non-linear kernel.

Model selection required tuning hyperparameters. Hyperparameters are
parameters which need to be specified or tuned by the user, in order to
train a model for a specific modeling problem. For example, when using
regularization, C is a hyperparameter that indicates the penalty for
overfitting. Hyperparameters are tuned using the training data to find
the best model. We selected hyperparameters by performing repeated
five-fold cross-validation (CV) on the training set {[}Figure 1{]}. The
five-fold CV was also stratified to maintain the overall case and
control distribution. We chose the hyperparameter values that led to the
best average CV predictive performance using the area under the receiver
operating characteristic curve (AUROC) {[}Figure S1 and S2{]}. The AUROC
ranges from 0, where the model's predictions are perfectly incorrect, to
1.0, where the model perfectly distinguishes between cases and controls.
An AUROC value of 0.5 indicates that model's predictions are no
different than random. To select hyperparameters, we performed a full
grid search for hyperparameter settings when training the models.
Default hyperparameter settings in previously developed ML packages
available in R, Python, and MATLAB programming languages may be
inadequate for effective application of classification algorithms and
need to be optimized for each new ML task. For example, L1-regularized
SVM with linear kernel showed large variability between different
regularization strengths (C) and needed to be tuned to find the optimal
C value {[}Figure S1{]}.

Once hyperparameters were selected, we trained the model using the full
training dataset and applied the final model to the held-out data to
evaluate the testing predictive performance of each model. The
data-split, hyperparameter selection, training and testing steps were
repeated 100 times to obtain a reliable and robust interpretation of
model performance {[}Figure 1{]}.

\textbf{Predictive performance and generalizability of the seven
models.} We evaluated the predictive performance of the seven models to
classify individuals as having normal colons or SRNs {[}Figure 2{]}. The
random forest model had higher test AUROC values than the other models
for detecting SRNs (resampling test between means of two groups, p
\textless{} 0.01). The median AUROC of the random forest model was 0.695
{[}IQR 0.650-0.739{]}. L2-regularized logistic regression, XGBoost,
L2-regularized SVM with linear and radial basis function kernel AUROC
values were not significantly different from one another and had median
AUROC values of 0.680 {[}IQR 0.639-0.750{]}, 0.679 {[}IQR
0.643-0.746{]}, 0.678 {[}IQR 0.639-0.750{]} and 0.668 {[}IQR
0.639-0.750{]}, respectively. L1-regularized SVM with linear kernel and
decision tree had significantly lower AUROC values than the other ML
models with median AUROC of 0.650 {[}IQR 0.629-0.760{]} and 0.601 {[}IQR
0.636-0.753{]}, respectively {[}Figure 2{]}. Interestingly, these
results demonstrate that the most complex model (XGBoost) did not have
the best performance and that the most interpretable models
(L2-regularized logistic regression and L2-regularized SVM with linear
kernel) performed nearly as well non-linear models. To support our
claim, we investigated the distribution of differences between random
forest and L2-regularized logistic regression AUROC values for each
data-split (Figure S3). Our results underscored that L2-regularized
logistic regression perform nearly as well as random forest.

@@

\textbf{Interpretation of each ML model.} Interpretability is related to
the degree to which humans can understand the reasons behind a model
prediction (32--34). Because we often use ML models not just to predict
a health outcome but also to identify the biomarkers for a disease,
model interpretation becomes crucial for microbiome studies. ML models
decrease in interpretability as they increase in complexity. In this
study we used two methods to help interpret our models.

First, we interpreted the feature importance of the linear models (L1
and L2-regularized SVM with linear kernel and L2-regularized logistic
regression) using the median rank of absolute feature weights for each
OTU {[}Figure 3{]}. We also reviewed the signs of feature weights to
determine whether an OTU was associated with classifying a subject as
being healthy or having an SRN. It was encouraging that many of the
highest ranked OTUs were shared across these three models, (e.g.~OTUs
50, 426, 609, 822, 1239). The benefit of this approach was that the
results of the analysis were based on the trained model parameters and
provided information regarding the sign and magnitude of the impact of
each OTU. However, this approach is only possible with linear models.

Second, to analyze non-linear models we interpreted the feature
importance using permutation importance. Whereas the absolute feature
weights were determined from the trained models, here we measured
importance using the held-out test data. Permutation importance analysis
is a post hoc explanation of the model, in which we randomly permuted
groups of perfectly correlated features together and other features
individually across the two groups in the held-out test data. We then
calculated how much the predictive performance of the model (i.e,
testing AUROC values) decreased when each OTU or group of OTUs was
randomly permuted. We ranked the OTUs based on how much the median
testing AUROC decreased when it was permuted; the OTU with the largest
decrease ranked highest {[}Figure 4{]}. Among the twenty OTUs with the
largest impact, there was only one OTU (OTU 822) that was shared among
all of the models; however, we found three OTUs (OTUs 58, 110, 367) that
were important in each of the tree-based models. Similarly, the random
forest and XGBoost models, shared four of the most important OTUs (OTUs
2, 12, 361, 477). Permutation analysis results also revealed that with
the exception of the decision tree model, removal of any individual OTU
had minimal impact on model performance. For example, if OTU 367 was
permuted across the samples in the decision tree model, the median AUROC
dropped from 0.601 to 0.525. In contrast, if the same OTU was permuted
in the random forest model, the AUROC only dropped from 0.695 to 0.680.
In this case, similar to previous studies (23, 35), it was not possible
to distinguish between health and disease using a single OTU. Although
permutation analysis allowed us to gauge the importance of an OTU, the
analysis was post hoc (i.e.~done using the test data) and these results
did not allow us to directly interrogate the models to know whether an
OTU was associated with classifying a subject as being healthy or having
an SRN.

To further highlight the differences between the two interpretation
methods, we used permutation importance to interpret the linear models
{[}Figure S5{]}. When we analyzed the L1-regularized SVM with linear
kernel model using feature rankings based on weights {[}Figure 3{]} and
permutation importance {[}Figure S5{]}, 17 of the 20 top OTUs (e.g.~OTU
609, 822, 1239) were deemed important by both interpretation methods.
Similarly, for the L2-regularized SVM and L2-regularized logistic
regression, 9 and 12 OTUs, respectively, were shared among the two
interpretation methods. Although permutation analysis does not not allow
us to determine the weight or the sign of the features, these results
indicate that both methods are consistent in selecting the most
important OTUs.

\textbf{The computational efficiency of each ML model.} We compared the
training times of the seven ML models. As expected, the training times
increased with the complexity of the model and the number of potential
hyperparameter combinations. Also, the linear models trained faster than
non-linear models {[}Figures S1-S2; Figure 5{]}.

@@

To address these needs, we developed an open-sourced framework to ML
models. Using this pipeline, we benchmarked seven ML models and showed
that the tradeoff between model complexity and performance may be less
severe than originally hypothesized. In terms of predictive performance,
the random forest model had the best AUROC compared to the other six
models. However, the second-best model was L2-regularized logistic
regression with a median AUROC difference less than 0.015 compared to
random forest. While our implementation of random forest took 83.2 hours
to train, our L2-regularized logistic regression trained in 12 minutes.
In terms of interpretability, random forest is a non-linear ML model and
is most often explained using post hoc methods such as permutation
importance. On the other hand, using L2-regularized logistic regression
we ranked the importance of each OTU based on their feature weights.
Comparing many different models showed us that the most complex model
was not necessarily the best model for our ML task.

We established a pipeline that can be generalized to any modeling method
that predicts a binary health outcome. We performed a random data-split
to create a training set (80\% of the data) and a held-out test set
(20\% of the data), which we used to evaluate predictive performance. We
repeated this data-split 100 times to measure the possible variation in
predictive performance. During the training, we tuned the model
hyperparameters with a repeated five-fold cross-validation. Despite the
high number of features microbiome datasets typically have, the models
we built with this pipeline were generalizable as shown by the good test
AUROCs.

We highlighted the importance of model interpretation to gain greater
biological insights into microbiota-associated diseases. In this study
we showcased two different interpretation methods: ranking each OTU by
(i) their absolute weights in the trained models and (ii) their impact
on the predictive performance based on permutation importance.
Human-associated microbial communities have complex correlation
structures which create collinearity in the datasets. This can hinder
our ability to reliably interpret models because the feature weights of
correlated OTUs are influenced by one another (36). To capture all
important features, once we identify highly ranked OTUs, we should
review their relationships with other OTUs. These relationships will
help us generate new hypotheses about the ecology of the disease and
test them with follow-up experiments. When we used permutation
importance, we took collinearity into consideration by grouping
correlated OTUs to determine their impact as a group. We grouped OTUs
that had a perfect correlation with each other, however, we could reduce
the correlation threshold to further investigate the relationships among
correlated features. It is important to know the correlation structures
of the data to avoid misinterpreting the models. This is likely to be a
particular problem with shotgun metagenomic datasets where collinearity
will be more pronounced due to many genes being correlated with one
another because they come from the same chromosome. To identify the true
underlying microbial factors of a disease, it is crucial to do
correlation analyses and further experimentation for biological
validation.

In this study, we did not consider all possible modeling approaches.
However, the principles highlighted throughout this study apply to other
ML modeling tasks with microbiome data. For example, we did not evaluate
multicategory classification methods to predict non-binary outcomes. We
could have trained models to differentiate between people with normal
colons and those with adenomas or carcinomas (k=3 categories). We did
not perform this analysis because the clinically relevant diagnosis
grouping was between patients with normal colons and those with SRNs.
Furthermore, as the number of classes increases, more samples are
required for each category to train an accurate model. We also did not
use regression-based analyses to predict a non-categorical outcome. We
have previously used such an approach to train random forest models to
predict fecal short-chain fatty acid concentrations based on microbiome
data (37). Our analysis was also limited to shallow learning methods and
did not explore deep learning methods such as neural networks. Deep
learning methods hold promise (12, 38, 39) but microbiome datasets often
suffer from having many features and small sample sizes, which result in
overfitting.

Our framework provides a reproducible structure to investigators wanting
to train, evaluate, and interpret their own ML models to generate
hypotheses regarding which OTUs might be biologically relevant. However,
deploying microbiome-based models to make clinical diagnoses or
predictions is a significantly harder and distinct undertaking. For
example, we currently lack standardized methods to collect patient
samples, generate sequence data, and report clinical data. We are also
challenged by the practical constraints of OTU-based approaches. The de
novo algorithms commonly in use are slow, require considerable memory,
and result in different OTU assignments as new data are added (40).
Finally, we also need independent validation cohorts to test the
performance of a diagnostic model. To realize the potential for using ML
approaches with microbiome data, it is necessary that we direct our
efforts to overcome these challenges.

Our study highlights the need to make educated choices at every step of
developing a ML model with microbiome data. We created an aspirational
rubric that researchers can use to identify potential pitfalls when
using ML in microbiome studies and ways to avoid them {[}Table S1{]}. We
have highlighted the trade-offs between model complexity and
interpretability, the need for tuning hyperparameters, the utility of
held-out test sets for evaluating predictive performance, and the
importance of considering correlation structures in datasets for
reliable interpretation. Furthermore, we underscored the importance of
proper experimental design and methods to help us achieve the level of
validity and accountability we want from models built for patient
health.

@@

\makecell[l]{SVM with \\radial basis kernel} & \makecell[l]{A classifier that is defined by an optimal Gaussian \\separating hyperplane that discriminates between labels.} & Non-linear \\ \hline

>>>>

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage[table]{xcolor}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{caption}
\usepackage{hyperref}
\usepackage{helvet} % Helvetica font
\renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
\usepackage[T1]{fontenc}
\usepackage[labelfont=bf]{caption}

@@

\(\dagger\) To whom correspondence should be addressed:
\href{mailto:pschloss@umich.edu}{\nolinkurl{pschloss@umich.edu}},
\href{mailto:wiensj@umich.edu}{\nolinkurl{wiensj@umich.edu}}

@@

Machine learning (ML) modeling of the human microbiome has the potential
to identify microbial biomarkers and aid in diagnosis of many diseases
such as inflammatory bowel disease, diabetes, and colorectal cancer.
Progress has been made towards developing ML models that predict health
outcomes using bacterial abundances, but inconsistent adoption of
training and evaluation methods call the validity of these models into
question. Furthermore, there appears to be a preference by many
researchers to favor increased model complexity over interpretability.
To overcome these challenges, we trained seven models that used fecal
16S rRNA sequence data to predict the presence of colonic screen
relevant neoplasias (SRNs; n=490 patients, 261 controls and 229 cases).
We developed a reusable open-source pipeline to train, validate and
interpret the models. To show the effect of model selection, we assessed
the predictive performance, interpretability, and training time of
L2-regularized logistic regression, L1 and L2-regularized support vector
machines (SVM) with linear and radial basis function kernels, decision
trees, random forest, and gradient boosted trees (XGBoost). The random
forest model performed best at detecting SRNs with an AUROC of 0.695
{[}IQR 0.651-0.739{]} but was slow to train (83.2 h) and not immidiately
interpretable. Despite its simplicity, L2-regularized logistic
regression followed random forest in predictive performance with an
AUROC of 0.680 {[}IQR 0.625-0.735{]}, trained faster (12 min), and
yielded a model that was inherently interpretable. Our analysis
highlights the importance of choosing an ML approach should be chosen
based on the goal of the study, as the choice will inform expectations
of performance and interpretability.

@@

As the number of people represented in human microbiome datasets grow,
there is an increasing desire to use microbiome data to diagnose
disease. However, the structure of the human microbiome is remarkably
variable among individuals to the point where it is often difficult to
identify the bacterial populations that are associated with diseases
using traditional statistical models. This variation is likely due to
the ability of many bacterial populations to fill the same niche such
that different populations cause the same disease in different
individuals. Furthermore, a growing number of studies have shown that it
is rare for a single bacterial species to be associated with a disease.
Instead, subsets of the microbiome account for differences in health.
Traditional statistical approaches do not adequately account for the
variation in the human microbiome and typically consider the protective
or risk effects of each bacterial population seperately (1). Recently,
machine learning (ML) models have grown in popularity among microbiome
researchers as our ability to to sample large numbers of individuals has
grown; such models can effectively account for the interpersonal
microbiome variation and the ecology of disease.

ML models can be used to increase our understanding of the variation in
the structure of existing data and in making predictions about new data.
Researchers have used ML models to diagnose and understand the
ecological basis of diseases such as liver cirrhosis, colorectal cancer,
inflammatory bowel diseases, obesity, and type 2 diabetes (2--19). The
task of diagnosing an individual relies on a rigorously validated model.
However, there are common methodological and reporting problems that
arise when applying ML to such data, that need to be addressed for the
field to progress. These problems include a lack of transparency in
which methods are used and how these methods are implemented; evaluating
models without a separate held-out test data; unreported variation
between the predictive performance on different folds of
cross-validation; and unreported variation between cross-validation and
testing performances. Though, the microbiome field is making progress to
avoid some of these pitfalls including validating their models on
independent datasets (8, 19, 20) and introducing ways to better use ML
tools (21--24), more work is needed to further improve reproducibility
and minimize overestimating for model performance.

Among microbiome researchers, the lack of justification when selecting a
modeling approach has often been due to an implicit assumption that more
complex models are better. This has resulted in a trend towards using
non-linear models such as random forest and deep neural networks (3, 12,
25--27) over simpler models such as logistic regression or other linear
models (19, 23, 28). Although in some cases complex models may capture
important non-linear relationships and therefore yield better
predictions, they can also result in black boxes that lack
interpretability. Such models require post hoc explanations to quantify
the importance of each feature in making predictions. Depending on the
goal of the modeling, other approaches may be more appropriate. For
example, researchers trying to identify the microbiota associated with
disease may desire a more interpretable model, whereas clinicians may
(but not always) emphasize predictive performance. Nonetheless, it is
important to understand that the benefit of more complex, less
interpretable models may be minimal (29, 30). It is important for
researchers to justify their choice of modeling approach.

To showcase a rigorous ML pipeline and to shed light on how ML model
selection can affect modeling results, we performed an empirical
analysis comparing seven modeling approaches with the same dataset and
pipeline. We built three linear models with different forms of
regularization: L2-regularized logistic regression and L1 and
L2-regularized support vector machines (SVM) with a linear kernel. We
also trained four non-linear models: SVM with radial basis function
kernel, a decision tree, random forest and gradient boosted trees. We
compared their predictive performance, interpretability, and training
time. To demonstrate the performance of these modeling approaches and
our pipeline, we used data from a previously published study that sought
to classify individuals as having normal colons or colonic lesions based
on the 16S rRNA gene sequences collected from fecal samples (4). This
dataset was selected because it is a relatively large collection of
individuals (N=490) connected to a clinically significant disease where
there is ample evidence that the disease is driven by variation in the
microbiome (2, 4, 5, 31). With this dataset, we developed an ML pipeline
that can be used in many different scenarios for training and evaluating
models. This framework can be easily applied to other host-associated
and environmental microbiome datasets.

@@

First, we randomly split the data into training and test sets so that
the training set consisted of 80\% of the full dataset, while the test
set was composed of the remaining 20\% {[}Figure 1{]}. To maintain the
distribution of controls and cases found in the full dataset, we
performed stratified splits. For example, our full dataset included 490
individuals. Of these, 261 had normal colons (53\%) and 229 had a screen
relevant neoplasia (SRN; 46.7\%). A training set included 393
individuals, of which 209 had an SRN (53\%), while the test set was
composed of 97 individuals of which 52 had an SRN (54\%). The training
data were used to build and select the models and the test set was used
for evaluating the model.

We trained seven different models using the training data {[}Table 1{]}.
We focused on different classification algorithms and regularization
methods. Regularization helps to prevent overfitting by penalizing a
model that fits the training data too well (32). For regularized
logistic regression and SVM with a linear kernel, we used
L2-regularization to keep all potentially important features. For
comparison, we also trained an L1-regularized SVM with a linear kernel.
L1-regularization on microbiome data led to a sparser solution (i.e.,
forced many coefficients to zero). To explore the potential for
non-linear relationships among features to improve classification, we
trained tree-based models including a decision tree, a random forest,
and gradient boosted trees (XGBoost) and an SVM with a non-linear
kernel.

Model selection requires tuning hyperparameters. Hyperparameters are
parameters that need to be specified or tuned by the user, in order to
train a model for a specific modeling problem. For example, when using
regularization, C is a hyperparameter that indicates the penalty for
overfitting. Hyperparameters are tuned using the training data to find
the best model. We selected hyperparameters by performing repeated
five-fold cross-validation (CV) on the training set {[}Figure 1{]}. The
five-fold CV was also stratified to maintain the overall case and
control distribution. We chose the hyperparameter values that led to the
best average CV predictive performance using the area under the receiver
operating characteristic curve (AUROC) {[}Figure S1 and S2{]}. The AUROC
ranges from 0, where the model's predictions are perfectly incorrect, to
1.0, where the model perfectly distinguishes between cases and controls.
An AUROC value of 0.5 indicates that model's predictions are no
different than random. To select hyperparameters, we performed a grid
search for hyperparameter settings when training the models. Default
hyperparameter settings in previously developed ML packages available in
R, Python, and MATLAB programming languages may be inadequate for
effective application of classification algorithms and should to be
optimized for each new ML task. For example, L1-regularized SVM with
linear kernel showed large variability between different regularization
strengths (C) and benefited from tuning {[}Figure S1{]}.

Once hyperparameters were selected, we trained the model using the full
training dataset and applied the final model to the held-out data to
evaluate the testing predictive performance of each model. The
data-split, hyperparameter selection, training and testing steps were
repeated 100 times to obtain a robust interpretation of model
performance, less likely to be affected by a ``lucky'' or ``unlucky''
split {[}Figure 1{]}.

\textbf{Predictive performance and generalizability of the seven
models.} We evaluated the predictive performance of the seven models to
classify individuals as having normal colons or SRNs {[}Figure 2{]}. The
random forest model had a median AUROC value of 0.695 {[}IQR
0.650-0.739{]} which was higher than the other models for detecting
SRNs. However, we determined that the random forest model performance
was not statistically significantly higher (p = 0.5) when we
investigated the distribution of differences between random forest and
L2-regularized logistic regression AUROC values for each data-split
(Figure S3). Similarly, L2-regularized logistic regression, XGBoost,
L2-regularized SVM with linear and radial basis function kernel AUROC
values were not significantly different from one another and had median
AUROC values of 0.680 {[}IQR 0.639-0.750{]}, 0.679 {[}IQR
0.643-0.746{]}, 0.678 {[}IQR 0.639-0.750{]} and 0.668 {[}IQR
0.639-0.750{]}, respectively. L1-regularized SVM with linear kernel and
decision tree had significantly lower AUROC values than the other ML
models with median AUROC of 0.650 {[}IQR 0.629-0.760{]} and 0.601 {[}IQR
0.636-0.753{]}, respectively {[}Figure 2{]}. Interestingly, these
results demonstrate that the most complex model (XGBoost) did not have
the best performance and that the most interpretable models
(L2-regularized logistic regression and L2-regularized SVM with linear
kernel) performed nearly as well non-linear models.

@@

\textbf{Interpretation of each ML model.} Interpretability is related to
the degree to which humans can understand the reasons behind a model
prediction (33--35). Because we often use ML models not just to predict
a health outcome, but also to identify potential biomarkers for disease,
model interpretation becomes crucial for microbiome studies. ML models
often decrease in interpretability as they increase in complexity. In
this study, we used two methods to help interpret our models.

First, we interpreted the feature importance of the linear models (L1
and L2-regularized SVM with linear kernel and L2-regularized logistic
regression) using the median rank of absolute feature weights for each
OTU {[}Figure 3{]}. We also reviewed the signs of feature weights to
determine whether an OTU was associated with classifying a subject as
being healthy or having an SRN. It was encouraging that many of the
highest ranked OTUs were shared across these three models, (e.g.~OTUs
50, 426, 609, 822, 1239). The benefit of this approach was that the
results of the analysis were based on the trained model parameters and
provided information regarding the sign and magnitude of the impact of
each OTU. However, this approach is limited to linear models or models
with prespecified interaction terms.

Second, to analyze non-linear models we interpreted the feature
importance using permutation importance (36). Whereas the absolute
feature weights were determined from the trained models, here we
measured importance using the held-out test data. Permutation importance
analysis is a post hoc explanation of the model, in which we randomly
permuted groups of perfectly correlated features together and other
features individually across the two groups in the held-out test data.
We then calculated how much the predictive performance of the model
(i.e, testing AUROC values) decreased when each OTU or group of OTUs was
randomly permuted. We ranked the OTUs based on how much the median
testing AUROC decreased when it was permuted; the OTU with the largest
decrease ranked highest {[}Figure 4{]}. Among the twenty OTUs with the
largest impact, there was only one OTU (OTU 822) that was shared among
all of the models; however, we found three OTUs (OTUs 58, 110, 367) that
were important in each of the tree-based models. Similarly, the random
forest and XGBoost models, shared four of the most important OTUs (OTUs
2, 12, 361, 477). Permutation analysis results also revealed that with
the exception of the decision tree model, removal of any individual OTU
had minimal impact on model performance. For example, if OTU 367 was
permuted across the samples in the decision tree model, the median AUROC
dropped from 0.601 to 0.525. In contrast, if the same OTU was permuted
in the random forest model, the AUROC only dropped from 0.695 to 0.680.
Although permutation analysis allowed us to gauge the importance of an
OTU, the analysis was post hoc (i.e., done using the test data).

To further highlight the differences between the two interpretation
methods, we used permutation importance to interpret the linear models
{[}Figure S5{]}. When we analyzed the L1-regularized SVM with linear
kernel model using feature rankings based on weights {[}Figure 3{]} and
permutation importance {[}Figure S5{]}, 17 of the 20 top OTUs (e.g.~OTU
609, 822, 1239) were deemed important by both interpretation methods.
Similarly, for the L2-regularized SVM and L2-regularized logistic
regression, 9 and 12 OTUs, respectively, were shared among the two
interpretation methods. These results indicate that both methods are
consistent in selecting the most important OTUs.

\textbf{The computational efficiency of each ML model.} We compared the
training times of the seven ML models. We did not take advantage of
parallelization in training nor was the code optimized for speed. Still,
as expected, the training times increased with the complexity of the
model and the number of potential hyperparameter combinations. Also, the
linear models trained faster than non-linear models {[}Figures S1-S2;
Figure 5{]}.

@@

To address these needs, we developed an open-sourced framework to ML
models. Using this pipeline, we benchmarked seven ML models and showed
that the tradeoff between model complexity and performance may be less
severe than originally hypothesized. In terms of predictive performance,
the random forest model had the best AUROC compared to the other six
models. However, the second-best model was L2-regularized logistic
regression with a median AUROC difference less than 0.015 compared to
random forest. While our implementation of random forest took 83.2 hours
to train, our L2-regularized logistic regression trained in 12 minutes.
In terms of interpretability, random forest is a non-linear ML model
while L2-regularized logistic regression, a linear model, is easily
interpreted according to the feature weights. Comparing many different
models showed us that the most complex model was not necessarily the
best model for our ML task.

We established a pipeline that can be generalized to any modeling method
that predicts a binary health outcome. We performed a random data-split
to create a training set (80\% of the data) and a held-out test set
(20\% of the data), which we used to evaluate predictive performance. We
repeated this data-split 100 times to measure the possible variation in
predictive performance. During training, we tuned the model
hyperparameters with a repeated five-fold cross-validation. Despite the
high number of features microbiome datasets typically have, the models
we built with this pipeline generalized to the held-out test sets.

We highlighted the importance of model interpretation to gain greater
biological insights into microbiota-associated diseases. In this study,
we showcased two different interpretation methods: ranking each OTU by
(i) their absolute weights in the trained models and (ii) their impact
on the predictive performance based on permutation importance.
Human-associated microbial communities have complex correlation
structures which create collinearity in the datasets. This can hinder
our ability to reliably interpret models because the feature weights of
correlated OTUs are influenced by one another (37). To capture all
important features, once we identify highly ranked OTUs, we should
review their relationships with other OTUs. These relationships will
help us generate new hypotheses about the ecology of the disease and
test them with follow-up experiments. When we used permutation
importance, we partially accounted for collinearity by grouping
correlated OTUs to determine their impact as a group. We grouped OTUs
that had a perfect correlation with each other, however, we could reduce
the correlation threshold to further investigate the relationships among
correlated features. It is important to know the correlation structures
of the data to avoid misinterpreting the models. This is likely to be a
particular problem with shotgun metagenomic datasets where collinearity
will be more pronounced due to many genes being correlated with one
another because they come from the same chromosome. To identify the true
underlying microbial factors of a disease, it is crucial to follow up
any correlation analyses with further hypothesis testing and
experimentation for biological validation.

In this study, we did not consider all possible modeling approaches.
However, the principles highlighted throughout this study apply to other
ML modeling tasks with microbiome data. For example, we did not evaluate
multicategory classification methods to predict non-binary outcomes. We
could have trained models to differentiate between people with normal
colons and those with adenomas or carcinomas (k=3 categories). We did
not perform this analysis because the clinically relevant diagnosis
grouping was between patients with normal colons and those with SRNs.
Furthermore, as the number of classes increases, more samples are
required for each category to train an accurate model. We also did not
use regression-based analyses to predict a non-categorical outcome. We
have previously used such an approach to train random forest models to
predict fecal short-chain fatty acid concentrations based on microbiome
data (38). Our analysis was also limited to shallow learning methods and
did not explore deep learning methods such as neural networks. Deep
learning methods hold promise (12, 39, 40) but microbiome datasets often
suffer from having many features and small sample sizes, which result in
overfitting.

Our framework provides a reproducible structure to investigators wanting
to train, evaluate, and interpret their own ML models to generate
hypotheses regarding which OTUs might be biologically relevant. However,
deploying microbiome-based models to make clinical diagnoses or
predictions is a significantly more challenging and distinct
undertaking. For example, we currently lack standardized methods to
collect patient samples, generate sequence data, and report clinical
data. We are also challenged by the practical constraints of OTU-based
approaches. The de novo algorithms commonly in use are slow, require
considerable memory, and result in different OTU assignments as new data
are added (41). Finally, we also need independent validation cohorts to
test the performance of a diagnostic model. To realize the potential for
using ML approaches with microbiome data, it is necessary that we direct
our efforts to overcome these challenges.

Our study highlights the need to make educated choices at every step of
developing an ML model with microbiome data. We created an aspirational
rubric that researchers can use to identify potential pitfalls when
using ML in microbiome studies and ways to avoid them {[}Table S1{]}. We
have highlighted the trade-offs between model complexity and
interpretability, the need for tuning hyperparameters, the utility of
held-out test sets for evaluating predictive performance, and the
importance of considering correlation structures in datasets for
reliable interpretation. Furthermore, we underscored the importance of
proper experimental design and methods to help us achieve the level of
validity and accountability we want from models built for patient
health.

@@

\makecell[l]{SVM with \\radial basis kernel} & \makecell[l]{A classifier that is defined by an optimal non-linear \\separating hyperplane that discriminates between labels.} & Non-linear \\ \hline
